%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
%\documentclass[sigplan,9pt,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For double-blind review submission, w/ CCS and ACM Reference
%\documentclass[sigplan,10pt,review,anonymous]{acmart}\settopmatter{printfolios=true}
% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
\documentclass[sigplan,9pt]{acmart}
\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For single-blind review submission, w/ CCS and ACM Reference
%\documentclass[sigplan,10pt,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission, w/ required CCS and ACM Reference
%\documentclass[sigplan,10pt]{acmart}\settopmatter{}


%% Conference information
%% Supplied to authors by publisher for camera-ready submission;
%% use defaults for review submission.
\acmConference[CoqPL'17]{CoqPL 2018 The Fourth International Workshop on Coq for Programming Languages}{January 13, 2018}{New York, NY, USA}
\acmYear{2018}
\acmISBN{} % \acmISBN{978-x-xxxx-xxxx-x/YY/MM}
\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}

%% Copyright information
%% Supplied to authors (based on authors' rights management selection;
%% see authors.acm.org) by publisher for camera-ready submission;
%% use 'none' for review submission.
\setcopyright{none}
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\copyrightyear{2017}           %% If different from \acmYear

%% Bibliography style
\bibliographystyle{ACM-Reference-Format}
%% Citation style
%\citestyle{acmauthoryear}  %% For author/year citations
%\citestyle{acmnumeric}     %% For numeric citations
%\setcitestyle{nosort}      %% With 'acmnumeric', to disable automatic
                            %% sorting of references within a single citation;
                            %% e.g., \cite{Smith99,Carpenter05,Baker12}
                            %% rendered as [14,5,2] rather than [2,5,14].
%\setcitesyle{nocompress}   %% With 'acmnumeric', to disable automatic
                            %% compression of sequential references within a
                            %% single citation;
                            %% e.g., \cite{Baker12,Baker14,Baker16}
                            %% rendered as [2,3,4] rather than [2-4].


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Note: Authors migrating a paper from traditional SIGPLAN
%% proceedings format to PACMPL format must update the
%% '\documentclass' and topmatter commands above; see
%% 'acmart-pacmpl-template.tex'.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


%% Some recommended packages.
\usepackage{booktabs}   %% For formal tables:
                        %% http://ctan.org/pkg/booktabs
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
                        %% http://ctan.org/pkg/subcaption
\include{preamble}

\begin{document}

%% Title information
\title{Linear Typing Judgments in Coq}         %% [Short Title] is optional;



%% Author information
%% Contents and number of authors suppressed with 'anonymous'.
%% Each author should be introduced by \author, followed by
%% \authornote (optional), \orcid (optional), \affiliation, and
%% \email.
%% An author may have multiple affiliations and/or emails; repeat the
%% appropriate command.
%% Many elements are not rendered, but should be provided for metadata
%% extraction tools.

%% Author with single affiliation.

\author{Jennifer Paykin}
\affiliation{
  \institution{University of Pennsylvania}            %% \institution is required
}
\email{jpaykin@seas.upenn.edu}          %% \email is recommended

\author{Robert Rand}
\affiliation{
  \institution{University of Pennsylvania}            %% \institution is required
}
\email{rrand@seas.upenn.edu}          %% \email is recommended

\author{Steve Zdancewic}
\affiliation{
  \institution{University of Pennsylvania}            %% \institution is required
}
\email{stevez@cis.upenn.edu}          %% \email is recommended

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
  Stuff
\end{abstract}


%% 2012 ACM Computing Classification System (CSS) concepts
%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
%\begin{CCSXML}
%<ccs2012>
%<concept>
%<concept_id>10011007.10011006.10011008</concept_id>
%<concept_desc>Software and its engineering~General programming languages</concept_desc>
%<concept_significance>500</concept_significance>
%</concept>
%<concept>
%<concept_id>10003456.10003457.10003521.10003525</concept_id>
%<concept_desc>Social and professional topics~History of programming languages</concept_desc>
%<concept_significance>300</concept_significance>
%</concept>
%</ccs2012>
%\end{CCSXML}

%\ccsdesc[500]{Software and its engineering~General programming languages}
%\ccsdesc[300]{Social and professional topics~History of programming languages}
%% End of generated code


%% Keywords
%% comma separated list
%\keywords{formal verification, type theory, quantum computing}  %% \keywords are mandatory in final camera-ready submission


%% \maketitle
%% Note: \maketitle command must come after title commands, author
%% commands, abstract environment, Computing Classification System
%% environment and commands, and keywords command.
\maketitle

\section{Introduction}

% Linear type systems have successfully been used in the past thirty years to
% safely program with a variety of effects including mutable state, memory
% management, concurrency, and quantum computing. 

\jennifer{The terminology is really bad in this section}

In the literature on linear type systems, two main perspectives have emerged. In
the ``kind'' perspective, data has either a linear or a persistent type, and
linear data can only be used once in a program.
The alternative is the ``usage'' perspective, where data is characterized by how
often is used is a program, and that usage is reflected in the types. These two
perspectives. These two presentations are equally expressive, and can be useful
in different domains. While the ``kind'' perspective is the traditional
presentation, and perhaps more natural in some cases, the ``usage'' perspective has
the advantage of being easy to implement, so there is significantly more tool
support for type systems built on this foundation. 

In this work we present tools to make it easier to implement traditional
``kind''-based linear type systems in Coq, using techniques that should extend
to other proof environments as well.

Consider a simple kind-based linear lambda calculus:
\[
    \inferrule*
    {~}
    {x:τ ⊢ x : τ}
\quad
    \inferrule*
    {Γ ∙ x:σ ⊢ e : τ}
    {Γ ⊢ λx.e : σ ⊸ τ}
\quad
    \inferrule*
    {Γ_1 ⊢ e_1 : σ ⊸ τ \\ Γ_2 ⊢ e_2 : σ}
    {Γ_1 ∙ Γ_2 ⊢ e_1 \; e_2 : τ}
\]

There are a few important differences from the simply-typed lambda calculus,
stemming from the fact that linear variables can neither be discarded or
duplicated.\footnote{We could also consider other substructural type systems,
  which consider different subsets of these rules, and although the details
  differ, many of the same principles apply.} First, in the variable rule, the
typing context must be exactly the singleton context $x:τ$; the corresponding
STLC rule, where $Γ ⊢ x : τ$ provided $(x:τ) ∈ Γ$, might discard additional
variables in $Γ$, contradicting linearity. Second, in the abstraction and
application rules, we write $Γ_1 ∙ Γ_2$ for the \emph{partial, disjoint merge}
of $Γ_1$ and $Γ_2$, which is undefined if $Γ_1 ∩ Γ_2 ≠ ∅$. This ensures that
linear variables cannot be duplicated to be used on both sides of an
application, or shadowed in the abstraction rule.

Unfortunately, type checking this judgment is not so straightforward. To show
that $Γ ⊢ e_1 \; e_2 : τ$, the typing context $Γ$ must be ``split'' in two disjoint
parts, depending on which variables occur in which parts. The ordinary
approach to linear typing judgments is to thread usage information through the
typing judgment. For example, one might instead consider a typing judgment
$Γ_{\text{in}} ⊢ e : τ > Γ_{\text{out}}$ where $e$ uses exactly the variables in
$Γ_{\text{in}} \textbackslash Γ_{\text{out}}$, meaning that $Γ_{\text{in}}$ is
thought of as ``input'' to the judgment, and $Γ_{\text{out}}$ is thought of as
``output''. This approach works well for \emph{affine} type systems (where linear
variables may be discarded but not duplicated) but to enforce true linearity,
one must consider a top-level judgment $Γ ⊢ e : τ > ∅$ \robert{hard to understand. 
Maybe just say  $Γ_{\text{out}}$ must be empty?}. In addition, this kind
of typing structure is not particularly elegant for additive types \todo{cite}\robert{maybe describe?},
and requires serializing the typing judgment. For example, to type an
application $e_1 \; e_2$, should it really matter if we type $e_1$ or $e_2$ first?

On the whole, we would prefer a judgment in Coq corresponding to the natural
presentation of linear types: $Γ ⊢ e : τ$, asserting that $e$ uses
\emph{exactly} the variables in $Γ$. Our goal in this work is to define a linear
typing judgment \coqe{Has_LType : Ctx -> Tm -> Ty -> Prop} in Coq and be able to
easily (automatically) discharge constraints stemming from the linearity of
typing contexts. In addition, the technique should not rely on a particular
representation of variable binding or typing contexts.

\section{Partiality}

Partial functions in Coq may take a number of forms. A partial function might be
represented as an inductive relation, but this has the disadvantage that the result
cannot be computed on concrete inputs. On the other hand, a partial function
might be represented as a function to the option type, which has the
disadvantage of obscuring compositionality, since composition must be surrounded
by the monadic bind. We can address this by lifting our partial function to options 
themselves, creating a total function on the domain plus $\bot$. We use this approach 
to write our disjoint merge operation.
\robert{A bit better? Note that there isn't a transition - it's not immediately clear why we're talking about partial functions.}
%A third option is to lift the partiality to all data
%values, so that disjoint merge is total, but may produce an invalid or undefined
%typing context as a result. This third option is the approach we take in this paper.
%\jennifer{I don't think this paragraph is super clear}

Thus, we can define the linear typing judgment in Coq as follows:
\todo{add to coq development}
\begin{coq}
Inductive Has_LType : Ctx -> Tm -> Ty -> Prop :=
| L_Var : ∀ Γ x τ, Γ = singleton x τ ->
                   Has_LType Γ (Var x) τ
| L_Abs : ∀ Γ x e σ, is_valid (Γ ∙ singleton x σ) ->
                     Has_LType (Γ ∙ singleton x σ) e τ ->
                     Has_LType Γ (Abs x e) (σ ⊸ τ)
| L_App : ∀ Γ1 Γ2 Γ e1 e2 σ τ, Γ1 ∙ Γ2 = Γ -> is_valid Γ ->
                               Has_LType Γ1 e1 (σ ⊸ τ) ->
                               Has_LType Γ2 e2 σ ->
                               Has_LType Γ (App e1 e2) τ.
\end{coq}

Consider trying to type the program $λx.λy.y x$. In Coq, the goal is expressed
as follows:
\begin{coq}
    forall x y σ τ, x ≠ y -> 
    Has_LType ∅ (Abs x (Abs y (App (Var y) (Var x)))) 
                (σ ⊸ (σ ⊸ τ) ⊸ τ)
\end{coq} 
Applying the abstraction constructors in sequence, we obtain the following goals:
\begin{coq}
  1) is_valid (∅ ∙ singleton x σ)
  2) is_valid (∅ ∙ singleton x σ ∙ singleton y (σ ⊸ τ))
  3) Has_LType (∅ ∙ singleton x σ ∙ singleton y (σ ⊸ τ)) 
               (App (Var y) (Var x)) τ
\end{coq}
We discuss how to solve goals of the form \coqe{is_valid Γ} in
\cref{sec:validity}. To progress with goal 3, we run into the problem described
in the introduction: how can an automatic type-checker guess the values of
\coqe{Γ1} and \coqe{Γ2} (and, for that matter, the value of $σ$)? 

Coq's existential variables (\coqe{evars}) provide a solution here; we defer the
instantiation of \coqe{Γ1} and \coqe{Γ2} to later, and use the tactic
\coqe{eapply L_App}. This introduces four new goals:
\begin{coq}
  4) ?Γ1 ∙ ?Γ2 = (∅ ∙  singleton x σ ∙ singleton y (σ ⊸ τ))
  5) is_valid (∅ ∙ singleton x σ ∙ singleton y (σ ⊸ τ))
  6) Has_LType ?Γ1 (Var y) (?σ0 ⊸ τ)
  7) Has_LType ?Γ2 (Var x) ?σ0
\end{coq}
Clearly, goals 6 and 7 can be discharged using the \coqe{LVar} constructor,
which resolves the \coqe{Evars} we introduced, and we are left with a revised
goal 4: 
\begin{coq}
  4) singleton y (σ ⊸ τ) ∙ singleton x σ 
  = ∅ ∙ singleton x σ ∙ singleton y (σ ⊸ τ)
\end{coq}

The typing rules thus introduce two new kinds of constraints: showing that two
contexts are equal, and showing that a context is valid.

\section{Equality of typing contexts}
\label{sec:monoid}

\robert{Using bottom but calling it zero seems confusing to me. Why not just use
  0 and 1?}
\jennifer{I'm worried about ∅ vs 0, but I think what you're suggesting is
  probably better, just have to rename some things}
The equality of typing contexts can be described as a partial commutative
monoid---that is, a commutative monoid $(⊤,∙)$ along with a zero element $⊥$
such that for all $a$, $a ∙ ⊥ = ⊥$. We characterize this structure as a type
class:
\begin{coq}
Class PCM A :=
    { one  : A
    ; zero : A
    ; m    : A -> A -> A }.
Notation "⊥" := zero.
Notation "⊤" := one.
Notation "a ∙ b" := (m a b) (left associativity, at level 20).
Class PCM_Laws A `{PCM A} :=
  { M_unit  : forall a, a ∙ ⊤ = a
  ; M_assoc : forall a b c, a ∙ (b ∙ c) = (a ∙ b) ∙ c
  ; M_comm  : forall a b, a ∙ b = b ∙ a
  ; M_absorb : forall a, a ∙ ⊥ = ⊥ 
  }.
\end{coq}

\citet{chlipala2013} describes a series of tactics for solving monoidal
equations using reflection, which we adapt to the partial commutative monoid
case. First, an LTac tactic reifies a term as a list of base variables, ignoring
units \coqe{⊤}, flattening out the associativity of $∙$, and collapsing to $⊥$
if it occurs anywhere in the term
\begin{coq}
Ltac reify a :=
  match a with
  | ⊤ => constr:(Some [])
  | ⊥ => constr:(None)
  | ?a1 ∙ ?a2 => let e1 := reify a1 in
                 let e2 := reify a2 in
                 match e1, e2 with
                 | Some ls1, Some ls2 => constr:(Some (ls1 ++ ls2))
                 | _, _ => constr:(None)
                 end
  | _ => constr:(Some [a])
  end.
\end{coq}
A goal \coqe{a1 = a2} can then be exchanged with one of the form %
\coqe{from_list ls1 = from_list ls2} where \coqe{lsi} is the result of calling
\coqe{reify} on \coqe{ai}. If both sides successfully resolve to lists, it
suffices to check whether the two are permutations of each other.

We can also deal with evars in the list, which is useful when considering
higher-order abstract syntax representations of variable binding.
\robert{Explain? Isn't this useful for instantiating unknown contexts?}


\section{Validity of typing contexts}
\label{sec:validity}

To prove statements of the form \coqe{is_valid Γ}, we introduce a boolean-valued
function \coqe{validity Γ}, and define \coqe{is_valid Γ} to be %
\coqe{validity Γ = true}. We then define a number of axioms describing the
behavior of validity with respect to the element of the partial monoid.
\begin{coq}
Class TypingContext (X A Ctx : Type) :=
  { singleton : X -> A -> Ctx; validity : Ctx -> bool }.
Class TypingContext_Laws X A Ctx `{PCM_Ctx : PCM_Laws Ctx} 
                                 `{TypingContext X A} :=
  { validity3 : forall Γ1 Γ2 Γ3, 
    validity (Γ1 ∙ Γ2 ∙ Γ3) =  validity (Γ1 ∙ Γ2) 
                            && validity (Γ1 ∙ Γ3) 
                            && validity (Γ2 ∙ Γ3)
  ; validity_reflection : forall Γ, validity Γ = false ↔ Γ = ⊥
  ; validity_singleton : forall x a, validity (singleton x a) = true
  ; validity_top : validity ⊤ = true
  ; validity_singleton_merge : forall x y a b, 
    validity (singleton x a ∙ singleton y b) = (x ≠? y) }. 
\end{coq}
The \coqe{validity3} axiom says that the merge of three typing contexts is
valid if and only if every binary combination is valid. The
\coqe{validity_reflection} axiom says that the only invalid context is the
undefined context. The next two axioms say that singleton contexts and the empty
context are both valid, and the final axiom says that the merge of two singleton
contexts is valid if and only if their domains are disjoint. The \coqe{validate}
tactic simply rewrites the target by these lemmas. For example, showing the
validity of goal (4) above, that
\begin{coq}
    is_valid (∅ ∙ singleton x σ ∙ singleton y (σ ⊸ τ))
\end{coq}
reduces to showing that \coqe{x ≠? y = true}, which we can conclude from the
hypothesis that \coqe{x ≠ y}.




% %% Acknowledgments
% \begin{acks}                            %% acks environment is optional
%                                         %% contents suppressed with 'anonymous'
%   %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
%   %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
%   %% acknowledge financial support and will be used by metadata
%   %% extraction tools.
%   This material is based upon work supported by the
%   \grantsponsor{GS100000001}{National Science
%     Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
%   No.~\grantnum{GS100000001}{nnnnnnn} and Grant
%   No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
%   conclusions or recommendations expressed in this material are those
%   of the author and do not necessarily reflect the views of the
%   National Science Foundation.
% \end{acks}


%% Bibliography
\bibliography{bibliography}


\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
